High geomagnetic field intensity recorded by anorthosite xenoliths requires a strongly powered late Mesoproterozoic geodynamo

### 2022-05-06 PNAS_manuscript/PNAS_manuscript.tex
Updates from Overleaf
Implement edits according to USGS comments

%\begin{figure*}
\begin{SCfigure*}
\centering
\noindent\includegraphics[width=12.4 cm]{Petro_QDM_full.png}
\caption{\footnotesize{Thin section petrographic images (A,B,C,D) and magnetic maps (E,F,G,H) of an anorthosite sample from the Beaver River anorthosite xenolith in the Silver Bay region from which paleomagnetic site AX16 and geochronology sample MS99033 were collected (left column;  \citealp{Zhang2021b}) and from a distinct anorthosite within the Duluth Complex Anorthositic Series (right column). The Duluth Complex anorthosites were not targeted for paleointensity experiments in this study given complexities associated with more pronounced fabrics. Cross-polarized petrographic images of the Beaver River anorthosite (A,C) reveal plagioclase with a granoblastic texture of crystals that are largely free of large opaque inclusions. In contrast, plagioclase crystals in the Duluth Complex anorthosite have euhedral, interlocking crystals with an igneous foliation (B) and the plagioclase crystals contain abundant Fe-Ti oxide needles that have preferred orientations that are often parallel with the [001] axis of the plagioclase. Maps of the vertical component of magnetic field (B$_z$) developed with a quantum diamond microscope (QDM) show relatively weak magnetic sources within plagioclase crystals in the Beaver River anorthosite (in E) relative to the strongly magnetic large oxide needles within Duluth Complex plagioclase (in F). The B$_z$ color scale is an order of magnitude greater in the Duluth Complex anorthosite maps (F, H) than the maps for the Beaver River anorthosite (E, G). Panels (E) to (G) and (F) to (H) show experiments performed on both samples where we apply a first field along the y axis of the field of view and then apply a second field of 300 mT orthogonal to the first field direction. The remanent magnetizations acquired by both anorthosites were mapped after the application of each field. The magnetic images show that remanent magnetizations of Beaver River anorthosite (i.e. the individual dipoles visible with paired red +B$_z$ and blue -B$_z$ lobes) align well with the first applied field direction (E) and then rotated to align with the second applied field direction indicating minimal anisotropic behavior (G). In contrast, the remanent magnetizations carried by the magnetic needles in plagioclase 1 of the Duluth anorthosite xenolith align well with the first applied field but those in plagioclase 2 acquired an oblique remanence direction with respect to the field direction (F). After the application of an orthogonal 300 mT external field, magnetization of those needles in plagioclase 1 did not change direction due to strong shape anisotropy whereas the magnetization of needles in plagioclase 2 flipped, but with the acquired remanence remaining oblique to the field direction (H). Insets in (E) to (H) show example dipole directions (fit using the algorithm of \cite{Lima2016a}) in response to the application of orthogonal IRMs. The directional changes of the Beaver River anorthosite xenolith sources indicate a lower magnetic anisotropy compared to the relative lack of change for the Duluth Complex anorthosite xenolith sources. The unit of the axes in the QDM maps are in $\mu$m.}}
\label{fig:Petro_QDM}
%\end{figure*}
\end{SCfigure*}

@@

\begin{figure*}
\noindent\includegraphics[width=17.8cm]{IZZI_examples.pdf}
\centering
\caption{\footnotesize{Example results of paleointensity experiments are displayed on Arai plots and zero-field heating results are shown as inset orthogonal plots (Zijderveld plots) for anorthosite and diabase specimens. Red (blue) circles indicate zero-field/in-field (in-field/zero-field) steps `ZI’ (`IZ’). Triangles mark partial thermal remanent magnetization (pTRM) checks. Blue and red squares in the Zijderveld plots are X–Y and X–Z projections, respectively, of the NRMs in specimen coordinates. Dashed red and blue lines show bounding regions associated with the \textit{SCAT} statistic \cite{Shaar2013a}. Plots on the top row show successful specimen paleointensity results with straight, single-slope behaviors that pass our selection criteria. The green lines represent fits for the dominant single-slope component that passes the acceptance criteria and gives an estimate of the ancient field strength (B$_{anc}$). The plots for anorthosite specimens AX1-1a and diabase BD5-1b on the bottom row show non-ideal sagging behaviour that fails our acceptance criteria. Specimen BD2-5a is an example where the data appear linear with distinct slopes in the low and high temperature ranges such that it could pass less restrictive selection criteria, particularly if a narrower temperature range was used for the experiment. Data analysis and visualization was conducted using PmagPy \cite{Tauxe2016a}.}}
\label{fig:IZZI_examples}
\end{figure*}

@@

\subsection*{Rock magnetism}
\subsubsection*{Coercivity}
Additional rock magnetic data indicate that those anorthosite specimens which pass the paleointensity selection criteria contain dominant magnetic remanence carriers with magnetic properties similar to stochiometric, non-interacting, single domain magnetite, whereas anorthosite samples that failed the paleointensity result selection along with all diabase samples have more pronounced populations of non-ideal carriers. Magnetic property measurement system (MPMS) data show that both the diabase and anorthosite contain (titano)magnetite as evidenced by the presence of the Verwey transition (Fig. S2;  \citealp{Verwey1939a, Feinberg2015a}). Anorthosite specimens from sites that yield successful paleointensity results have Verwey transition temperatures near 120 K as expected for stochiometric magnetite with minimal Ti contents \cite{Ozdemir1993a}. However, diabase and anorthosite specimens that did not pass our paleointensity selection typically have Verwey transitions that are suppressed toward lower temperatures (Fig. S2), indicating that their magnetic grains either have relatively higher Ti content or have been partially oxidized \cite{Ozdemir1993a}. Another difference is that samples which pass paleointensity selection have distinctly higher average median destructive field values than other anorthosite and diabase specimens (Fig. \ref{fig:coercivity}). Single-component fits for coercivity spectra \cite{Maxbauer2016a} show that anorthosites having successful paleointensity results can contain magnetic grain populations that have higher coercivities. For these successful samples, median destructive destructive (MDF) field values associated with back-field demagnetization experiments range from 44 to 144 mT with a median of 53 mT (Fig. \ref{fig:coercivity}). In contrast, other anorthosite and diabase specimens tend to have lower peak coercivities (MDF range of 18-35 mT with a median of 22 mT; Fig. \ref{fig:coercivity}). This result is consistent with an interpretation that magnetic grain populations with more multidomain-like behavior are responsible for the non-ideal paleointensity behaviors during experiments of such specimens \cite{Xu2004a}.

\begin{figure}
\noindent\includegraphics[width=\linewidth]{coercivity.pdf}
\centering
\caption{\footnotesize{Top: Example coercivity spectra of anorthosite and diabase chips from sites that pass or fail our paleointensity selection criteria. Bottom: Box plots of median destructive field (MDF) values associated with back-field demagnetization for all anorthosite and diabase specimens with single-component coercivity unmixing results. Both plots show that anorthosite specimens that pass paleointensity selection criteria have higher coercivities consistent with a higher portion of single-domain-like magnetite grains than the other anorthosite specimens and the diabase.}}
\label{fig:coercivity}
\end{figure}

@@

\subsection*{Considering secular variation and cooling rate}
To best characterize the geomagnetic axial dipole field intensity during a certain time period a paleointensity dataset should cover a sufficient amount of time such that paleosecular variations of the geomagnetic field are averaged. Thermal modeling results from ref. \citealp{Zhang2021b} indicate that the Beaver River anorthosite xenoliths were heated to tholeiitic magma temperature ($\sim$1100\textdegree C)---lower than the melting temperature of the anorthosite plagioclase (given its composition of 70\% anorthite; \citealp{Morrison1983a, Doyle2016a}) and acquired thermal remanent magnetization during cooling with their diabase host on a time scale of a few thousand years, partially averaging secular variation within single sites. Another consequence of the anorthosite xenoliths having slowly cooled in the interior of thick diabase intrusions is that slow cooling rates can bias paleointensity estimates toward higher values \cite{Halgedahl1980a}. Large differences in cooling rates between acquisition of an NRM in nature versus a TRM in the lab can result in overestimated paleointensities for single domain grains \cite{Dodson1980a, Halgedahl1980a, Nagy2021a}. From the thermal history model of \cite{Zhang2021b}, we can estimate the duration over which the diabase and anorthosite cooled from the Curie temperature of magnetite ($\sim$580\textdegree C) to the time when they blocked in the majority of their characteristic natural remanence magnetization ($\sim$500\textdegree  C; Fig. \ref{fig:IZZI_examples}; \citealp{Zhang2021b}). We find the cooling time to $\sim$500\textdegree C to be $\sim$1.5 kyr, which corresponds to a cooling rate of $\sim1.7\times10^{-9}$ $^\circ$C s$^{-1}$. In contrast, the lab cooling rate is much faster through the same temperature interval with an estimated cooling rate of $\sim1.3\times10^{-1}$ $^\circ$C s$^{-1}$. The significant cooling rate difference leads to a predicted $\sim$33\% overestimate of true ancient field following the model of \citealp{Halgedahl1980a} (Fig. S3). This estimate on cooling rate effect is similar to the value of $\sim$30\% overestimate derived from the model of ref. \citealp{Nagy2021a}. We therefore correct our paleointensity results by a factor of 0.75. Remanence held by vortex state (pseudo-single domain) and multidomain grains is not as biased by cooling rate \cite{Biggin2013a}. The potential for some remanence to be held by these grains, as suggested by slightly zigzagging Arai plots (Fig. \ref{fig:IZZI_examples}), could mean that this factor is an over-correction and true paleointensity values could be slightly higher than those reported here. The cooling rate-corrected specimen paleointensity estimates together with specimen- and site-level means are shown in Fig. \ref{fig:PINT_cooling_corrected}.

Anorthosite xenoliths have paleointensity results that are consistent within small regions, but vary between regions. Anorthosite AX12 (28.65 $\pm$ 4.40 $\mu$T) and AX16 (33.20 $\pm$ 3.82 $\mu$T) in the Silver Bay area were emplaced $\sim$450 meters apart and have indistinguishable paleointensity estimates (Figs. \ref{fig:Geologic_map} and \ref{fig:PINT_cooling_corrected}). $\sim$10 km to the north, anorthosite xenoliths AX11 (55.14 $\pm$ 0.05 $\mu$T) and AX13 (57.62 $\pm$ 3.20 $\mu$T) in the Kennedy Creek area were also emplaced closely ($\sim$125 meters apart) and yield similar values to one another, but distinct paleointensity estimates from those of AX12 and AX16 (Figs. \ref{fig:Geologic_map} and \ref{fig:PINT_cooling_corrected}). $\sim$30 km further to the north in the Carlton Peak region, AX6 (22.80 $\pm$ 11.41 $\mu$T), AX8 (39.40 $\pm$ 9.14 $\mu$T), and AX10 (35.20 $\pm$ 0.83 $\mu$T) are anorthosite xenoliths within 10 meters of one another and also yield similar paleointensity estimates albeit with relatively large uncertainties (Fig. \ref{fig:PINT_cooling_corrected}). The anorthosites of the three distinct regions captured three different intervals of geomagnetic field intensities during the emplacement and cooling of the Beaver River diabase sills ca. 1092 Ma. To contextualize the high paleointensity values, we plot the regional mean values (Table. S3) based on specimen-level paleointensity results together with calculated paleointensity values at 22\textdegree N latitude based on the PADM2M model of the time-varying geomagnetic field over the past 2 million years \cite{Ziegler2011a} in Figure \ref{fig:PINT_cooling_corrected}B. The locality-mean values of these three regions are plotted in Figure \ref{fig:PINT_cooling_corrected}C (and reported in Table. S3) where we also present the distribution of means calculated from 3 random values taken from the PADM2M model for a total of 10000 iterations. In addition, we perform the same comparison with the existing paleointensity data for the Cenozoic Era (the last 66 million years) from the PINT database (filtered by Q$_{PI}\geq$3; Fig. S5) given that the observational dataset could record more geomagnetic field excursions or regional high flux patches. The anorthosite xenoliths three region-mean intensity is higher than all values from the resample results of the PADM2M model, and rival the top 3\% values of the resample averages from the Cenozoic data (Fig. S5). This comparison supports that the anorthosite xenoliths record an exceptionally strong geomagnetic field ca. 1092 Ma.

\begin{figure*}[h!]
\noindent\includegraphics[width=17.8 cm]{PINT_compilation.pdf}
\centering
\caption{\footnotesize{Compilation of calculated virtual (axial) dipole moment values from the PINT database (PINT v8.0.0; \url{http://www.pintdb.org/}; \citealp{Bono2021a}), including all Phanerozoic VDM and VADM records with Q$_{PI}$ values $>$3 and additional Neoproterozoic data from refs. \citealp{Lloyd2021a, Lloyd2021b, Thallner2021a, Thallner2021b}. Paleointensity estimates from refs. \citealp{Pesonen1983a} and \citealp{Kulakov2013a} are not included in the compilation due to the specimen-level double-slope behavior as discussed in the text. Overall, the anorthosite xenoliths from this study record a high Mesoproterozoic field exceeding the value projected by the second order polynomial curve from ref. \citealp{Bono2019a} which is based on an interpretation of there being a monotonic decay of the geodynamo through the Proterozoic. The highest site-mean virtual dipole moment of the anorthosites would rank in the top 2$\%$ of those in the database for the Cenozoic Era (the last 66 million years) when there was unequivocally a crystallizing inner core. The y axis maximum is set at the 99 percentile of the compiled Cenozoic paleointensity data.}}
\label{fig:PINT_compilation}
\end{figure*}

@@

The hypothesis that a weak Ediacaran geomagnetic field is a telltale sign of the lack of an inner core with core nucleation following shortly thereafter may predict that it is the most significant weak to strong field transition in the paleointensity record. However, Fig. \ref{fig:PINT_compilation} shows that transitions from low to high field intensities occurred before, during, and after the Ediacaran Period. In the Ediacaran record developed to date, there is a two-fold increase in Earth's virtual dipole moment when comparing estimates from the ca. 565 Ma Sept-$\hat{I}$les intrusions \cite{Bono2019a} to those from ca. 550 Ma volcanics of the Skinner Cove Formation (Fig. \ref{fig:PINT_compilation}; \citealp{Thallner2021a}). In the late Mesoproterozoic, there is at least a six-fold increase within a period of $\sim$35 Myr from a low average virtual dipole moment of $\sim$13 ZAm$^2$ recorded by the ca. 1140 Ma Abitibi dikes which yielded straight Arai plots \cite{Macouin2003a}, to a high moment of $\sim$70 ZAm$^2$ recorded by the ca. 1106 Ma Osler Volcanics, with even stronger values from ca. 1092 Ma by Beaver River anorthosite xenoliths that record virtual dipole moments up to $\sim$129 ZAm$^2$. While the ca. 1140 Abitibi dikes' paleointensity estimates are not as low as those of the ca. 565 Ma Sept-$\hat{I}$les intrusions, this virtual dipole moment increase in the Mesoproterozoic from the ca. 1140 Ma data to the ca. 1100 Ma data is the largest yet documented in the Precambrian on a 10s of millions of years timescale (Fig. \ref{fig:PINT_compilation}). The tempo and scale of this field intensity transition could match with model-based predictions associated with the onset of inner core nucleation \cite{Davies2021a}. This ca. 1.1 Ga timing would be broadly consistent with the ca. 1.3 Ga onset proposed by ref. \citealp{Biggin2015a} albeit later given the exclusion of previous overestimated paleointensity values from the ca. 1.3 Ga Gardar basalts that are superseded by data in ref. \citealp{Kodama2019a}. However, a model prediction of sustained strong field values following inner core nucleation is challenged by data from the ca. 1070 Ma Bangemall Sills which include a sill with a low virtual dipole moment of $\sim$6.4 ZAm$^2$ (Fig. \ref{fig:PINT_compilation};  \citealp{Lloyd2021b}). Following the Ediacaran, there are also low paleointensity estimates from Devonian rocks such as the ca. 370 Ma dikes and lavas of the Siberian Viluy Traps that give virtual dipole moment estimates of 4.3 to 14.9 ZAm$^2$ (Fig. \ref{fig:PINT_compilation}; \citealp{Hawkins2019a}). These low values as well as data from the ca. 414 Ma Strathmore lava flows \cite{Hawkins2021a}, the ca. 410-380 Ma lava flows of Siberia and the Kola Peninsula \cite{Shcherbakova2017a}, the ca. 408–393 Ma Buribay volcanics \cite{Shcherbakova2021a} and the ca. 332 Ma Kinghorn volcanics \cite{Hawkins2021a} has led to the proposal of this interval being the ``Mid-Paleozoic Dipole Low'' \cite{Hawkins2021a}. This ``Mid-Paleozoic Dipole Low'' is followed by high paleointensity values such that there is a six-fold increase in the database from virtual dipole moment of $\sim$16 ZAm$^2$ ca. 332 Ma to $\sim$99 ZAm$^2$ ca. 308 Ma in the late Carboniferous (Fig. \ref{fig:PINT_compilation}; \citealp{Hawkins2021a}).

>>>>

%\begin{figure*}
\begin{SCfigure*}
\centering
\noindent\includegraphics[width=12.4 cm]{Petro_QDM_full.pdf}
\caption{\footnotesize{Thin section petrographic images (A,B,C,D) and magnetic maps (E,F,G,H) of an anorthosite sample from the Beaver River anorthosite xenolith in the Silver Bay region from which paleomagnetic site AX16 and geochronology sample MS99033 were collected (left column;  \citealp{Zhang2021b}) and from a distinct anorthosite within the Duluth Complex Anorthositic Series (right column). The Duluth Complex anorthosites were not targeted for paleointensity experiments in this study given complexities associated with more pronounced fabrics. Cross-polarized petrographic images of the Beaver River anorthosite (A,C) reveal plagioclase with a granoblastic texture of crystals that are largely free of large opaque inclusions. In contrast, plagioclase crystals in the Duluth Complex anorthosite have euhedral, interlocking crystals with an igneous foliation (B) and the plagioclase crystals contain abundant Fe-Ti oxide needles that have preferred orientations that are often parallel with the [001] axis of the plagioclase. Maps of the vertical component of magnetic field (B$_z$) developed with a quantum diamond microscope (QDM) show relatively weak magnetic sources within plagioclase crystals in the Beaver River anorthosite (in E) relative to the strongly magnetic large oxide needles within Duluth Complex plagioclase (in F). The B$_z$ color scale is an order of magnitude greater in the Duluth Complex anorthosite maps (F, H) than the maps for the Beaver River anorthosite (E, G). Panels (E) to (G) and (F) to (H) show experiments performed on both samples where we apply a first field along the y-axis of the field of view and then apply a second field of 300 mT orthogonal to the first field direction. The remanent magnetizations acquired by both anorthosites were mapped after the application of each field. The magnetic images show that remanent magnetizations of Beaver River anorthosite (i.e. the individual dipoles visible with paired red +B$_z$ and blue -B$_z$ lobes) align well with the first applied field direction (E) and then rotated to align with the second applied field direction indicating minimal anisotropic behavior (G). In contrast, the remanent magnetizations carried by the magnetic needles in plagioclase 1 of the Duluth anorthosite xenolith align well with the first applied field but those in plagioclase 2 acquired an oblique remanence direction with respect to the field direction (F). After the application of an orthogonal 300 mT external field, magnetization of those needles in plagioclase 1 did not change direction due to strong shape anisotropy whereas the magnetization of needles in plagioclase 2 flipped, but with the acquired remanence remaining oblique to the field direction (H). Insets in (E) to (H) show example dipole directions (fit using the algorithm of \cite{Lima2016a}) in response to the application of orthogonal IRMs. The directional changes of the Beaver River anorthosite xenolith sources indicate a lower magnetic anisotropy compared to the relative lack of change for the Duluth Complex anorthosite xenolith sources. The unit of the axes in the QDM maps are in $\mu$m.}}
\label{fig:Petro_QDM}
%\end{figure*}
\end{SCfigure*}

@@

\begin{figure*}
\noindent\includegraphics[width=17.8cm]{IZZI_examples.pdf}
\centering
\caption{\footnotesize{Example results of paleointensity experiments are displayed on Arai plots and zero-field heating results are shown as inset orthogonal plots (Zijderveld plots) for anorthosite and diabase specimens. Green squares and lines in Arai plots and green squares in orthogonal plots show the range of data points used for fitting. Red (blue) circles indicate zero-field/in-field (in-field/zero-field) steps `ZI’ (`IZ’). Triangles mark partial thermal remanent magnetization (pTRM) checks. Blue and red squares in the Zijderveld plots are X–Y and X–Z projections, respectively, of the NRMs in specimen coordinates. Dashed red and blue lines show bounding regions associated with the \textit{SCAT} statistic \cite{Shaar2013a}. Plots on the top row show successful specimen paleointensity results with straight, single-slope behaviors that pass our selection criteria. The green lines represent fits for the dominant single-slope component that passes the acceptance criteria and gives an estimate of the ancient field strength (B$_{anc}$). The plots for anorthosite specimens AX1-1a and diabase BD5-1b on the bottom row show non-ideal sagging behaviour that fails our acceptance criteria. Specimen BD2-5a is an example where the data appear linear with distinct slopes in the low and high temperature ranges such that it could pass less restrictive selection criteria, particularly if a narrower temperature range was used for the experiment. Data analysis and visualization was conducted using PmagPy \cite{Tauxe2016a}.}}
\label{fig:IZZI_examples}
\end{figure*}

@@

\subsection*{Rock magnetism}
\subsubsection*{Coercivity}
Additional rock magnetic data indicate that those anorthosite specimens which pass the paleointensity selection criteria contain dominant magnetic remanence carriers with magnetic properties similar to stoichiometric, non-interacting, single domain magnetite, whereas anorthosite samples that failed the paleointensity result selection along with all diabase samples have more pronounced populations of non-ideal carriers. Magnetic property measurement system (MPMS) data show that both the diabase and anorthosite contain (titano)magnetite as evidenced by the presence of the Verwey transition (Fig. S2;  \citealp{Verwey1939a, Feinberg2015a}). Anorthosite specimens from sites that yield successful paleointensity results have Verwey transition temperatures near 120 K as expected for stochiometric magnetite with minimal Ti contents \cite{Ozdemir1993a}. However, diabase and anorthosite specimens that did not pass our paleointensity selection typically have Verwey transitions that are suppressed toward lower temperatures (Fig. S2), indicating that their magnetic grains either have relatively higher Ti content or have been partially oxidized \cite{Ozdemir1993a}. Another difference is that samples which pass paleointensity selection have distinctly higher average median destructive field values than other anorthosite and diabase specimens (Fig. \ref{fig:coercivity}). Single-component fits for coercivity spectra \cite{Maxbauer2016a} show that anorthosites having successful paleointensity results can contain magnetic grain populations that have higher coercivities. For these successful samples, median destructive (MDF) field values associated with back-field demagnetization experiments range from 44 to 144 mT with a median of 53 mT (Fig. \ref{fig:coercivity}). In contrast, other anorthosite and diabase specimens tend to have lower peak coercivities (MDF range of 18-35 mT with a median of 22 mT; Fig. \ref{fig:coercivity}). This result is consistent with an interpretation that magnetic grain populations with more multidomain-like behavior are responsible for the non-ideal paleointensity behaviors during experiments of such specimens \cite{Xu2004a}.

\begin{figure}
\noindent\includegraphics[width=\linewidth]{coercivity.pdf}
\centering
\caption{\footnotesize{Top: Example coercivity spectra of anorthosite and diabase chips from sites that pass or fail our paleointensity selection criteria. Bottom: Box plots of median destructive field (MDF) values associated with back-field demagnetization for all anorthosite and diabase specimens with single-component coercivity unmixing results. Bars in the boxes show median values and the whiskers show the minimum and the maximum value of the datasets. Both plots show that anorthosite specimens that pass paleointensity selection criteria have higher coercivities consistent with a higher portion of single-domain-like magnetite grains than the other anorthosite specimens and the diabase.}}
\label{fig:coercivity}
\end{figure}

@@

\subsection*{Considering secular variation and cooling rate}
To best characterize the geomagnetic axial dipole field intensity during a certain time period, a paleointensity dataset should cover a sufficient amount of time such that paleosecular variations of the geomagnetic field are averaged. Thermal modeling results from ref. \citealp{Zhang2021b} indicate that the Beaver River anorthosite xenoliths were heated to tholeiitic magma temperature ($\sim$1100\textdegree C)---lower than the melting temperature of the anorthosite plagioclase (given its composition of 70\% anorthite; \citealp{Morrison1983a, Doyle2016a}) and acquired thermal remanent magnetization during cooling with their diabase host on a time scale of a few thousand years, partially averaging secular variation within single sites. Another consequence of the anorthosite xenoliths having slowly cooled in the interior of thick diabase intrusions is that slow cooling rates can bias paleointensity estimates toward higher values \cite{Halgedahl1980a}. Large differences in cooling rates between acquisition of an NRM in nature versus a TRM in the lab can result in overestimated paleointensities for single domain grains \cite{Dodson1980a, Halgedahl1980a, Nagy2021a}. From the thermal history model of \cite{Zhang2021b}, we can estimate the duration over which the diabase and anorthosite cooled from the Curie temperature of magnetite ($\sim$580\textdegree C) to the time when they blocked in the majority of their characteristic natural remanence magnetization ($\sim$500\textdegree  C; Fig. \ref{fig:IZZI_examples}; \citealp{Zhang2021b}). We find the cooling time to $\sim$500\textdegree C to be $\sim$1.5 kyr, which corresponds to a cooling rate of $\sim1.7\times10^{-9}$ $^\circ$C s$^{-1}$. In contrast, the lab cooling rate is much faster through the same temperature interval with an estimated cooling rate of $\sim1.3\times10^{-1}$ $^\circ$C s$^{-1}$. The significant cooling rate difference leads to a predicted $\sim$33\% overestimate of true ancient field following the model of \citealp{Halgedahl1980a} (Fig. S3). This estimate on cooling rate effect is similar to the value of $\sim$30\% overestimate derived from the model of ref. \citealp{Nagy2021a}. We therefore correct our paleointensity results by a factor of 0.75. Remanence held by vortex state (pseudo-single domain) and multidomain grains is not as biased by cooling rate \cite{Biggin2013a}. The potential for some remanence to be held by these grains, as suggested by slightly zigzagging Arai plots (Fig. \ref{fig:IZZI_examples}), could mean that this factor is an over-correction and true paleointensity values could be slightly higher than those reported here. The cooling rate-corrected specimen paleointensity estimates together with specimen- and site-level means are shown in Fig. \ref{fig:PINT_cooling_corrected}.

Anorthosite xenoliths have paleointensity results that are consistent within small regions, but vary between regions. Anorthosite AX12 (28.65 $\pm$ 4.40 $\mu$T) and AX16 (33.20 $\pm$ 3.82 $\mu$T) in the Silver Bay area were emplaced $\sim$450 meters apart and have indistinguishable paleointensity estimates (Figs. \ref{fig:Geologic_map} and \ref{fig:PINT_cooling_corrected}). $\sim$10 km to the north, anorthosite xenoliths AX11 (55.14 $\pm$ 0.05 $\mu$T) and AX13 (57.62 $\pm$ 3.20 $\mu$T) in the Kennedy Creek area were also emplaced closely ($\sim$125 meters apart) and yield similar values to one another, but distinct paleointensity estimates from those of AX12 and AX16 (Figs. \ref{fig:Geologic_map} and \ref{fig:PINT_cooling_corrected}). $\sim$30 km further to the north in the Carlton Peak region, AX6 (22.80 $\pm$ 11.41 $\mu$T), AX8 (39.40 $\pm$ 9.14 $\mu$T), and AX10 (35.20 $\pm$ 0.83 $\mu$T) are anorthosite xenoliths within 10 meters of one another and also yield similar paleointensity estimates albeit with relatively large uncertainties (Fig. \ref{fig:PINT_cooling_corrected}). The anorthosites of the three distinct regions captured three different intervals of geomagnetic field intensities during the emplacement and cooling of the Beaver River diabase sills ca. 1092 Ma. To contextualize the high paleointensity values, we plot the regional mean values (Table S3) based on specimen-level paleointensity results together with calculated paleointensity values at 22\textdegree N latitude based on the PADM2M model of the time-varying geomagnetic field over the past 2 million years \cite{Ziegler2011a} in Figure \ref{fig:PINT_cooling_corrected}B. The locality-mean values of these three regions are plotted in Figure \ref{fig:PINT_cooling_corrected}C (and reported in Table S3) where we also present the distribution of means calculated from 3 random values taken from the PADM2M model for a total of 10000 iterations. In addition, we perform the same comparison with the existing paleointensity data for the Cenozoic Era (the last 66 million years) from the PINT database (filtered by Q$_{PI}\geq$3; Fig. S5) given that the observational dataset could record more geomagnetic field excursions or regional high flux patches. The anorthosite xenoliths three region-mean intensity is higher than all values from the resample results of the PADM2M model, and rival the top 3\% values of the resample averages from the Cenozoic data (Fig. S5). This comparison supports that the anorthosite xenoliths record an exceptionally strong geomagnetic field ca. 1092 Ma.

\begin{figure*}[h!]
\noindent\includegraphics[width=17.8 cm]{PINT_compilation.pdf}
\centering
\caption{\footnotesize{Compilation of calculated virtual (axial) dipole moment values from the PINT database (PINT v8.0.0; \url{http://www.pintdb.org/}; \citealp{Bono2021a}), including all Phanerozoic VDM and VADM records with Q$_{PI}$ values $>$3 and additional Neoproterozoic data from refs. \citealp{Lloyd2021a, Lloyd2021b, Thallner2021a, Thallner2021b}. Paleointensity estimates from refs. \citealp{Pesonen1983a} and \citealp{Kulakov2013a} are not included in the compilation due to the specimen-level double-slope behavior as discussed in the text. Overall, the anorthosite xenoliths from this study record a high Mesoproterozoic field exceeding the value projected by the second order polynomial curve from ref. \citealp{Bono2019a} which is based on an interpretation of there being a monotonic decay of the geodynamo through the Proterozoic. The highest site-mean virtual dipole moment of the anorthosites would rank in the top 2$\%$ of those in the database for the Cenozoic Era (the last 66 million years) when there was unequivocally a crystallizing inner core. The y-axis maximum is set at the 99th percentile of the compiled Cenozoic paleointensity data.}}
\label{fig:PINT_compilation}
\end{figure*}

@@

The hypothesis that a weak Ediacaran geomagnetic field is a telltale sign of the lack of an inner core with core nucleation following shortly thereafter may predict that it is the most significant weak to strong field transition in the paleointensity record. However, Fig. \ref{fig:PINT_compilation} shows that transitions from low to high field intensities occurred before, during, and after the Ediacaran Period. In the Ediacaran record developed to date, there is a two-fold increase in Earth's virtual dipole moment when comparing estimates from the ca. 565 Ma Sept-$\hat{I}$les intrusions \cite{Bono2019a} to those from ca. 550 Ma volcanics of the Skinner Cove Formation (Fig. \ref{fig:PINT_compilation}; \citealp{Thallner2021a}). In the late Mesoproterozoic, there is at least a six-fold increase within a period of $\sim$35 Myr from a low average virtual dipole moment of $\sim$13 ZAm$^2$ recorded by the ca. 1140 Ma Abitibi dikes which yielded straight Arai plots \cite{Macouin2003a}, to a high moment of $\sim$70 ZAm$^2$ recorded by the ca. 1106 Ma Osler Volcanics, with even stronger values from ca. 1092 Ma by Beaver River anorthosite xenoliths that record virtual dipole moments up to $\sim$129 ZAm$^2$. While the ca. 1140 Abitibi dikes' paleointensity estimates are not as low as those of the ca. 565 Ma Sept-$\hat{I}$les intrusions, this virtual dipole moment increase in the Mesoproterozoic from the ca. 1140 Ma data to the ca. 1100 Ma data is the largest yet documented in the Precambrian on a 10s of millions of years timescale (Fig. \ref{fig:PINT_compilation}). The tempo and scale of this field intensity transition could match with model-based predictions associated with the onset of inner core nucleation \cite{Davies2021a}. This ca. 1.1 Ga timing would be broadly consistent with the ca. 1.3 Ga onset proposed by ref. \citealp{Biggin2015a} albeit later given the exclusion of previous overestimated paleointensity values from the ca. 1.3 Ga Gardar basalts that are superseded by data in ref. \citealp{Kodama2019a}. However, a model prediction of sustained strong field values following inner core nucleation is challenged by data from the ca. 1070 Ma Bangemall Sills which include a sill with a low virtual dipole moment of $\sim$6.4 ZAm$^2$ (Fig. \ref{fig:PINT_compilation};  \citealp{Lloyd2021b}). Following the Ediacaran, there are also low paleointensity estimates from Devonian rocks such as the ca. 370 Ma dikes and lavas of the Siberian Viluy Traps that give virtual dipole moment estimates of 4.3 to 14.9 ZAm$^2$ (Fig. \ref{fig:PINT_compilation}; \citealp{Hawkins2019a}). These low values as well as data from the ca. 414 Ma Strathmore lava flows \cite{Hawkins2021a}, the ca. 410-380 Ma lava flows of Siberia and the Kola Peninsula \cite{Shcherbakova2017a}, the ca. 408–393 Ma Buribay volcanics \cite{Shcherbakova2021a} and the ca. 332 Ma Kinghorn volcanics \cite{Hawkins2021a} have led to the proposal of this interval being the ``Mid-Paleozoic Dipole Low'' \cite{Hawkins2021a}. This ``Mid-Paleozoic Dipole Low'' is followed by high paleointensity values such that there is a six-fold increase from virtual dipole moments of $\sim$16 ZAm$^2$ ca. 332 Ma to $\sim$99 ZAm$^2$ ca. 308 Ma in the late Carboniferous (Fig. \ref{fig:PINT_compilation}; \citealp{Hawkins2021a}).
